Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to the ones obtained using deep learning.
%\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  In this appliction it is not fesible to model interactions,
so the model is retricted to main effects.

%\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
We also investigate more complex supervised
approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
around genetic susceptibility to obesity that only five of the 97 genomic loci have an independent association signal
the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
similarly complex phenotype that requires higher order dependencies between features to be taken into
account. %, as done by RF and DL.

RF inherit many of the benefits of decision trees. It handles large numbers of variables transparently and
provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
population classes sizes). They provide a measure of variable importance and allow for interactions between
variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
is not feasible.  We note that decision trees (and random forests) model local variable interactions -- there are 
no main effects or global interaction effects.

>>>>

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to those obtained using deep learning.
%\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  In this appliction, it is not feasible to model interactions and
so the model is restricted to main effects.

%\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
We also investigate more complex supervised
approaches, namely Random Forest (RF), with the rationale being that interaction effects may dominate. A recent meta-study by
Locke {\it et al.}~\cite{Locke2015} found across 97 genomic loci associated with genetic susceptibility to obesity, only five
had an independent association signal and the rest seem to jointly predict this complex metabolic phenotype.  
We therefore investigate whether ethnicity is a similarly complex phenotype that requires higher order dependencies between
features to be taken into account. %, as done by RF and DL.

RF inherit many of the benefits of decision trees. It handles large numbers of variables transparently and
provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
population classes sizes). RF provide a measure of variable importance and allow for interactions between
variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
is not feasible.  We note that decision trees (and random forests) model local variable interactions -- there are 
no main effects or global interaction effects.
